Length Gauge Theory of Surface Second Harmonic Generation

### 2014-11-20 paper-v-0.tex
finished conclusions

In this section we present a 
relevant test case to check the
consistency of our approach. We have selected
a clean Si(001) surface with a $2\times 1$ surface reconstruction.
The slab for such a surface could be chosen to be centrosymmetric 
by creating the front and back
surfaces with the same $2\times 1$
reconstruction. However, we choose to terminate one 
of the surfaces with hydrogen producing an ideal  
terminated bulk Si surface. The H atoms simply saturate the 
dangling bonds of the bulk-like Si atoms at the surface, as seen in 
Fig.~\ref{si2x1}.
We take the $z$ coordinate pointing out of the surface and the $x$
coordinate, along the crystallographic [011] direction, is parallel to the dimmers. 
\begin{figure}
\centering 
\includegraphics[scale=.8]{images/si2x1-crop}
\caption{(color on line) The slab shows a 
    clean Si(001)$2\times 1$ front surface 
    with an ideal terminated Si bulk back surface. The
    dangling bonds are H (small balls) saturated. 
    This image depicts 12 Si atomic layers
    with one H atomic layer. 
\label{si2x1}} 
\end{figure} 
The idea behind this slab configuration is that the 
cristalline symmetry of the H terminated surface imposes that 
$\chi_{\mathrm{H}}^{xxx}=0$.
 The $2\times 1$ surface has no 
such restrictions, so $\chi_{2\times 1}^{xxx}\ne 0$.
This is due to the fact that along the $y$ direction there is a mirror plane for the
H-saturated surface, whereas 
for the $2\times 1$ surface this mirror is lost as the dimers are
asymmetric along $x$. 
Thus, calculating $\chi^{xxx}$ for the full-slab, or the 
half-slab containing the $2\times 1$ surface\cite{note1}
should yield the same result since the contribution from the H
saturated surface is zero regardless. 
We must check that the following 
relationship is satisfied for this particular slab
\begin{align}\label{hs}
\chi_{\mathrm{half-slab}}^{xxx}(-2\go;\go,\go) 
=
\chi_{\mathrm{full-slab}}^{xxx}(-2\go;\go,\go) 
,
\end{align}
where
$\chi_{\mathrm{half-slab}}^{xxx}(-2\go;\go,\go)$ is calculated using
$C(z)=1$ from the upper half containing the $2\times 1$ 
surface reconstruction, as seen in Fig.~\ref{si2x1},
and $\chi_{\mathrm{full-slab}}^{xxx}(-2\go;\go,\go)$ is calculated using
$C(z)=1$ through the full slab.
We show the results for this comparison in the remainder 
of this section

@@

In Fig.~\ref{fig2}
we compare 
$\chi^{xxx}_{\mathrm{half-slab}}$  
vs. 
$\chi^{xxx}_{\mathrm{full-slab}}$ 
for the four different possibilities 
between including or not including the
effects of $\bfv^\nl$ or the scissors correction
$\hbar\sigma$.   
For these results we chose
$\hbar\sigma=0.5$ eV, that is the GW gap reported in
Ref.~\onlinecite{rohlfingPRB95,garciaCPC01}. 
We see that for all four instances the 
difference between responses is quite small.
Indeed, when the value $|\chi^{xxx}|$ 
is large the difference between the two is very small; 
when the value is small the difference increases only slightly, 
but the spectra is so close to zero that these it is negligible. 
These differences would decrease as the number of atomic layers 
increases. We remark that 32 layers in the slab is more than enough 
to confirm that the extraction of the surface second harmonic 
susceptibility from the $2\times 1$ surface is readily possible 
using the formalism contained in Eq.~\eqref{chis}.
This confirms the validity of our theory and is the main result of
this article; through the proposed layer formalism we can calculate the surface SH
pseudopotentials
and the many-body effects through the scissors correction.
Our scheme ought to work for any slab.

@@

We proceed to explain 
some of the features seen in $|\chi^{xxx}_{2\times 1}|$ that, as
explained above, are obtained 
by calculating $|\chi^{xxx}_{\mathrm{half-slab}}|$.
First, from Fig.~\ref{fig2} we note a series of resonances 
that derive from 1$\omega$ and 2$\omega$ terms in
Eq.~\eqref{chis}. 
Notice that the 
2$\omega$ resonances start below $E_g/2$ where $E_g=??$ eV is the LDA
band gap of this surface, and $E_g=??$ eV is the scissor shifted gap for
$\hbar\sigma=0.5$ eV.
These resonances
come from the electronic states of the 
$2\times 1$ surface, that lie inside the bulk band gap of Si and are the 
well known electronic surface states.
In Fig.~\ref{fig3}
We see that the effect of $\bfv^\nl$ reduces the value of   
$|\chi^{xxx}_{2\times 1}|$,
without changing the 
spectral shape; this also can be confirmed from
the cases of zero scissors correction from Fig.~\ref{fig2}.
%
From Fig.~\ref{fig4}
we note that the scissors correction 
shifts the spectra to higher energies as expected. However, contrary 
to the case of linear optics\cite{cabellosPRB09} the shift introduced
by the scissors correction is not 
rigid, as pointed out in Ref.~\onlinecite{nastosPRB05}.
 This is because the second harmonic optical response mixes 
$1\omega$ and $2\omega$ transitions (see Eq.~\eqref{chis}), and accounts 
for the non-rigid shift.
\begin{figure}
\centering 
\includegraphics[scale=.8]{plots/fig3}
\caption{(color on line) 
$\chi^{xxx}_{2\times 1}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer, 
with and without the contribution from $\bfv^\nl$.
\label{fig3}} 
\end{figure}
\begin{figure}
\centering 
\includegraphics[scale=.8]{plots/fig4}
\caption{(color on line) 
$\chi^{xxx}_{2\times 1}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer, 
for two different values of 
the scissors correction $\hbar\sigma$. 
\label{fig4}} 
\end{figure}
Also, when we compare 
$|\chi^{xxx}_{2\times 1}|$ for two values of $\hbar\sigma$,
we see that the first two peaks are almost rigidly 
shifted with a small difference in height while
the rest of the peaks are modified substantially. 
This behavior comes from the fact that the first two
peaks are almost exclusively related to the 
2$\omega$ resonances of
Eq.~\eqref{chis}. The other peaks are a combination 
of 1$\omega$ and 2$\omega$ resonances 
and yield a more varied spectrum. 
This way we see that small changes in the value of the 
scissors shift can in general affect the SSH susceptibility 
spectrum quite dramatically.
We mention that
the value of $\hbar\sigma=0.5$ eV, used above, 
is the ``average'' GW gap taken from 
Ref.~\onlinecite{rohlfingPRB95} 
that is in agreement with Ref.~\onlinecite{garciaCPC01}. 
The value of $\hbar\sigma=0.63$ eV is the ``average'' 
gap taken from Refs.~\onlinecite{asahiPRB00}, 
where more $\bfk$-points in the Brillouin zone were
used to calculate its GW value.

Although, local fields are neglected, we expect that parallel to an
interface they should be in principle 
small, as the electric field is continuous. So, we would expect that the $xxx$ component 
of 
$\bfgchi(-2\go;\go,\go)$, would have an small influence from the local fields.
Unfortunately the experimental measurement of the $xxx$
component of $\bfgchi(-2\go;\go,\go)$ is not possible, as the SH 
radiated intensity would be proportional not only to this component 
but also to other components of $\bfgchi(-2\go;\go,\go)$,  
however,
in a forthcoming publication 
we will present a study of SSHG from several 
Si surfaces 
with comparison to experimental 
results.

@@

We have presented a formulation to calculate the surface second harmonic
(SSH) susceptibility tensor $\bfgchi(-2\go;\go,\go)$, using the length gauge
formalism and within the independent particle approximation (IPA).
It includes on the same footing and for the first time in the literature 
i) the scissors correction, ii) the cut function, and iii) the 
We have used a Si(001)$2\times 1$ surface to confirm that our scheme
correctly obtains the surface response as we confirm
that 
$\chi_{\mathrm{half-slab}}^{xxx}(-2\go;\go,\go) 
\approx
\chi_{\mathrm{full-slab}}^{xxx}(-2\go;\go,\go) 
. 
$
Although, one can in principle increase the number of atomic layers,
$\bfk$-points, etc.  to
improve even further on the similarity of the half-slab and full-slab results, we
hava chosen a good compromise between accuracy and the burden and time
of the computational scheme. 
We describe the effect of the independent inclusion of the three
effects mentioned above in the calculation of  
$\bfgchi(-2\go;\go,\go)$. To wit,
the scissors correction shifts the spectrum to higher energies, though
the shifting is not rigid and mixes the 1$\omega$ and 2$\omega$
resonances, the cut functions
allow us
to extract unequivocally   
$\chi^{xxx}_{2\times 1}(-2\go;\go,\go)$, 
and the effects of the nonlocal part of the pseudopotentials,
reduces the value of   
$|\chi^{xxx}_{2\times 1}(-2\go;\go,\go)|$, keeping the same lineshape. 
The $xxx$ component of  
$\bfgchi_{2\times 1}(-2\go;\go,\go)$, 
could not be
experimentally  isolated,
however in a forthcoming publication 
we will compare our formulation 
against
experimental 
results. 
We have neglected
local field 
and excitonic effects.
Although these are important factors in the optical response of a semiconductor,
their efficient calculation is theoretically and 
numerically challenging and still 
under debate.\cite{beyond}  
This merits further study but is beyond the scope of this paper. 
Nevertheless, the inclusion of aforementioned contributions 
in
our scheme opens the unprecedented possibility to study 
surface SHG with more versatility and providing 
more accurate results than previous approaches.

>>>>

In this section we present a 
relevant test case to check the
consistency of our approach. We have selected
a clean Si(001) surface with a $2\times 1$ surface reconstruction.
The slab for such a surface could be chosen to be centrosymmetric 
by creating the front and back
surfaces with the same $2\times 1$
reconstruction. However, we choose to terminate one 
of the surfaces with hydrogen producing an ideal  
terminated bulk Si surface. The H atoms simply saturate the 
dangling bonds of the bulk-like Si atoms at the surface, as seen in 
Fig.~\ref{si2x1}.
We take the $z$ coordinate pointing out of the surface and the $x$
coordinate along the crystallographic [011] direction is parallel to the dimmers. 
\begin{figure}
\centering 
\includegraphics[scale=.8]{images/si2x1-crop}
\caption{(color on line) The slab shows a 
    clean Si(001)$2\times 1$ front surface 
    with an ideal terminated Si bulk back surface. The
    dangling bonds are H (small balls) saturated. 
    This image depicts 12 Si atomic layers
    with one H atomic layer. 
\label{si2x1}} 
\end{figure} 
The idea behind this slab configuration is that the 
cristalline symmetry of the H terminated surface imposes that 
$\chi_{\mathrm{H}}^{xxx}=0$.
 The $2\times 1$ surface has no 
such restrictions, so $\chi_{2\times 1}^{xxx}\ne 0$.
This is due to the fact that along the $y$ direction there is a mirror plane for the
H-saturated surface, whereas 
for the $2\times 1$ surface this mirror is lost as the dimers are
asymmetric along $x$. 
Thus, calculating $\chi^{xxx}$ for the full-slab, or the 
half-slab containing the $2\times 1$ surface\cite{note1}
should yield the same result since the contribution from the H
saturated surface is zero regardless. 
We must check that the following 
relationship is satisfied for this particular slab
\begin{align}\label{hs}
\chi_{\mathrm{half-slab}}^{xxx}(-2\go;\go,\go) 
=
\chi_{\mathrm{full-slab}}^{xxx}(-2\go;\go,\go) 
,
\end{align}
where
$\chi_{\mathrm{half-slab}}^{xxx}(-2\go;\go,\go)$ is calculated using
$C(z)=1$ from the upper half containing the $2\times 1$ 
surface reconstruction, as seen in Fig.~\ref{si2x1},
and $\chi_{\mathrm{full-slab}}^{xxx}(-2\go;\go,\go)$ is calculated using
$C(z)=1$ through the full slab.
We show the results for this comparison in the remainder 
of this section.

@@

In Fig.~\ref{fig2}
we compare 
$\chi^{xxx}_{\mathrm{half-slab}}$  
vs. 
$\chi^{xxx}_{\mathrm{full-slab}}$ 
for the four different possibilities 
between including or not including the
effects of $\bfv^\nl$ or the scissors correction
$\hbar\sigma$.   
For these results we chose
$\hbar\sigma=0.5$ eV, that is the GW gap reported in
Refs.~\onlinecite{rohlfingPRB95,garciaCPC01}. 
We see that for all four instances the 
difference between responses is quite small.
Indeed, when the value $|\chi^{xxx}|$ 
is large the difference between the two is very small; 
when the value is small the difference increases only slightly, 
but the spectra is so close to zero that these it is negligible. 
These differences would decrease as the number of atomic layers 
increases. We remark that 32 layers in the slab is more than enough 
to confirm that the extraction of the surface second harmonic 
susceptibility from the $2\times 1$ surface is readily possible 
using the formalism contained in Eq.~\eqref{chis}.
This confirms the validity of our theory and is the main result of
this article; through the proposed layer formalism we can calculate the surface SH
pseudopotentials
and the many-body effects through the scissors correction.
Our scheme should to work for any slab.

@@

We proceed to explain 
some of the features seen in $|\chi^{xxx}_{2\times 1}|$ that, as
explained above, are obtained 
by calculating $|\chi^{xxx}_{\mathrm{half-slab}}|$.
First, from Fig.~\ref{fig2} we note a series of resonances 
that derive from 1$\omega$ and 2$\omega$ terms in
Eq.~\eqref{chis}. 
Notice that the 
2$\omega$ resonances start below $E_g/2$ where $E_g=0.53$ eV is the LDA
band gap of this surface, and $E_g=1.035$ eV is the scissor shifted gap for
$\hbar\sigma=0.5$ eV.
These resonances
come from the electronic states of the 
$2\times 1$ surface, that lie inside the bulk band gap of Si and are the 
well known electronic surface states.
In Fig.~\ref{fig3}
We see that the effect of $\bfv^\nl$ reduces the value of   
$|\chi^{xxx}_{2\times 1}|$,
without changing the 
spectral shape; this also can be confirmed from
the cases of zero scissors correction from Fig.~\ref{fig2}.
%
From Fig.~\ref{fig4}
we note that the scissors correction 
shifts the spectra to higher energies as expected. However, contrary 
to the case of linear optics\cite{cabellosPRB09} the shift introduced
by the scissors correction is not 
rigid, as pointed out in Ref.~\onlinecite{nastosPRB05}.
 This is because the second harmonic optical response mixes 
$1\omega$ and $2\omega$ transitions (see Eq.~\eqref{chis}), and accounts 
for the non-rigid shift.
\begin{figure}
\centering 
\includegraphics[scale=.8]{plots/fig3}
\caption{(color on line) 
$\chi^{xxx}_{2\times 1}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer, 
with and without the contribution from $\bfv^\nl$.
\label{fig3}} 
\end{figure}
\begin{figure}
\centering 
\includegraphics[scale=.8]{plots/fig4}
\caption{(color on line) 
$\chi^{xxx}_{2\times 1}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer, 
for two different values of 
the scissors correction $\hbar\sigma$. 
\label{fig4}} 
\end{figure}
Also, when we compare 
$|\chi^{xxx}_{2\times 1}|$ for two values of $\hbar\sigma$,
we see that the first two peaks are almost rigidly 
shifted with a small difference in height while
the rest of the peaks are modified substantially. 
This behavior comes from the fact that the first two
peaks are almost exclusively related to the 
2$\omega$ resonances of
Eq.~\eqref{chis}. The other peaks are a combination 
of 1$\omega$ and 2$\omega$ resonances 
and yield a more varied spectrum. 
This way we see that small changes in the value of the 
scissors shift can in general affect the SSH susceptibility 
spectrum quite dramatically.
We mention that
the value of $\hbar\sigma=0.5$ eV, used above, 
is the ``average'' GW gap taken from 
Ref.~\onlinecite{rohlfingPRB95} 
that is in agreement with Ref.~\onlinecite{garciaCPC01}. 
The value of $\hbar\sigma=0.63$ eV is the ``average'' 
gap taken from Ref.~\onlinecite{asahiPRB00}, 
where more $\bfk$-points in the Brillouin zone were
used to calculate its GW value.

Although local fields are neglected they should, in principle, 
be small parallel to the interface as the electric field is continuous.
So, we would expect that the $xxx$ component of 
$\bfgchi(-2\go;\go,\go)$ would have a small influence from the local fields.
Unfortunately the experimental measurement of the $xxx$
component of $\bfgchi(-2\go;\go,\go)$ is not possible as the SH 
radiated intensity would be proportional not only to this component 
but also to the other components of $\bfgchi(-2\go;\go,\go)$.  
However,
in a forthcoming publication 
we will present a study of SSHG from several 
Si surfaces 
with comparison to experimental 
results.

@@

We have presented a formulation to calculate the surface second harmonic
(SSH) susceptibility tensor $\bfgchi(-2\go;\go,\go)$, using the length gauge
formalism and within the independent particle approximation (IPA).
It includes on equal footing and for the first time in the literature: 
i) the scissors correction, ii) the cut function, and iii) the
contribution of the non-local part of the pseudopotentials.
We have used a Si(001)$2\times 1$ surface to confirm that our scheme
correctly obtains the surface response as we confirm
that 
$\chi_{\mathrm{half-slab}}^{xxx}(-2\go;\go,\go) 
\approx
\chi_{\mathrm{full-slab}}^{xxx}(-2\go;\go,\go) 
. 
$
Although one can in principle increase the number of atomic layers,
$\bfk$-points, etc. to
improve even further on the similarity of the half-slab and full-slab results, we
hava chosen a good compromise between accuracy and the burden and time
of the computations. 
We describe the effect of the independent inclusion of the three
effects mentioned above in the calculation of  
$\bfgchi(-2\go;\go,\go)$. The scissors correction shifts the spectrum to 
higher energies though the shifting is not rigid and mixes the 
$1\omega$ and $2\omega$ resonances, the cut function allows us
to extract unequivocally $\chi^{xxx}_{2\times 1}(-2\go;\go,\go)$, 
and the effects of the nonlocal part of the pseudopotentials
reduces the value of   
$|\chi^{xxx}_{2\times 1}(-2\go;\go,\go)|$ keeping the same lineshape. 
The $xxx$ component of  
$\bfgchi_{2\times 1}(-2\go;\go,\go)$, 
can not be
experimentally isolated,
however in a forthcoming publication 
we will compare our formulation 
against
experimental 
results. 
We have neglected
local field 
and excitonic effects.
Although these are important factors in the optical response of a semiconductor,
their efficient calculation is theoretically and 
numerically challenging and still 
under debate.\cite{beyond}  
This merits further study but is beyond the scope of this paper. 
Nevertheless, the inclusion of aforementioned contributions 
in
our scheme opens the unprecedented possibility to study 
surface SHG with more versatility and provides 
more accurate results than previous approaches.
